---
title: "Emergency Services Call Analysis in Buffalo, New York"
author: Sydney Deck
subtitle: Exploring the Correlation Between Mental Health–Related Emergency Calls and Median Household Income.
date: today
date-format: long
editor: 
  markdown: 
    wrap: 72
code-fold: true
---

# 1. Introduction

Emergency services are essential in times of crisies, providing
necessary response teams to areas in need with support from Police,
Fire, and Medical professionals. Unfortunately high call volumes and low
staffing can result in misguided directives and improper handling of
sensitive situations. This report will provide further insight on call
classification and distribution of mental health related calls within
the City of Buffalo, including any potential correlation between median
household income. The results of this analysis will assist in
understanding the use of Buffalo's resources and how socio-economic
factors can be indicative of improper resource distribution and improved
community development.

# 2. Materials and Methods

Overview: Import, clean and categorize required data from March and
April of 2024 into Mental Health categories. Plot call types and call
volumes. Run a regression analysis on specified mental health call types
and median income values.

Data required for this analysis: - 911 call data provided by the City of
Buffalo Police Department for the months of March and April in 2024.
Data can be found at
'https://figshare.com/articles/dataset/Buffalo_911_Call_March_and_April_2024/30585986?file=59454170' -
Census tract data from 'https://www.census.gov/data/' using an API key
from 2023, the most recent and compatible dataset. Must include median
income data and total populaton data. - Zipcode boundaries from
'https://data.buffalony.gov/' using an API key.

Necessary steps: 1. Import necessary libraries and datasets for use. 2.
Clean up imported data and complete pre-processing requirements. 3.
Define mental health categories and assign them to the call data. 4.
Create a summary table of call types and locations. 5. Plot a bar chart
summary of call frequency 6. Map most frequent call types by zipcode 7.
Map call volume by zip code 8. Create and run a regression model on call
frequency versus median income. 9. Plot regression.

### 2.1 Data Sources

```{r, message=F, warning=F}
#| results: hide
#install.packages(c("tigris", "sf", "plotly", "broom"))
#install.packages("tidycensus", dependencies = TRUE)
```

First, necessary libraries for data processing, retrieval and
organization will be downloaded.

```{r, message=F, warning=F}
#| results: hide
#| 
# Data processing libraries
library(dplyr)
library(tidyr)
library(readr)

# Mapping & Spatial Data 
library(sf)
library(leaflet)
library(htmltools)

# Census & Geographical Data Retrieval 
library(tigris)
library(tidycensus)

# Visualization
library(ggplot2)
library(plotly)

# Modeling and Statistical Output
library(broom)
```

With these libraries installed, the 911 data can be imported from a CSV
file saved within the GitHub data folder. Zipcode data can be imported
by using an API through the OpenData Buffalo website. A similar process
is used for the census data, using an API key and the TIGRIS library
previously installed.

```{r, message=F, warning=F}
#| results: hide
#| 
# Import 911 data CSV from github data folder
data_911_url <- "https://raw.githubusercontent.com/sjdeck/geo511/refs/heads/master/data/mar2apr2024_latlong.csv"
data_911 <- read_csv(data_911_url)

# Import zipcode data from API 
url_json <- "https://data.buffalony.gov/resource/qnyw-efar.geojson"
zipcode_json <- st_read(url_json, quiet = TRUE)

# Import Census data through API 

# Set API key
census_api_key("e12a911e4853b9b6ae9944d643816d70cd2b9b18", install = FALSE)
readRenviron("~/.Renviron") 

# Download census data
census_data <- get_acs(
  geography = "zcta",
 variables = c(
    median_income = "B19013_001",  
    total_pop     = "B01003_001"),          
  year = 2023,
  survey = "acs5",
  geometry = FALSE
)

```

### 2.2 Preprocessing and Data Preparation

##### Cleaning and Filtering 911 records

Now with all the data imported, data preprocessing measures are taken to
ensure smooth transitioning and analysis. The original 911 data is not
suitable for spatial analysis, so converting to an SF object will help
with mapping and spatial analysis. Projection transformation on the
zipcode and census data is also required to match the 911 data.

```{r, message=F, warning=F}

# Convert to an sf object to "enable standard spatial data handling, analysis, and visualization"
calls_sf <- st_as_sf(data_911, coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)

# Transform zipcode data to match 911 data
zipcode_json <- st_transform(zipcode_json, st_crs(calls_sf))

# Join call data and zipcode data together 
calls_sf <- st_intersection(calls_sf, zipcode_json %>% select(zcta5ce10))

# Clean census data, change CRS and pivot wider
census_clean <- census_data %>%
  select(GEOID, variable, estimate, GEOID) %>%
  pivot_wider(names_from = variable, values_from = estimate)
```

##### Categorizing and assigning call descriptions

Paid Partnership for Buffalo requested a mental health call analysis for
a potential pilot program, assisting areas with high mental health calls
with resources other than police officers during emergencies that may
not require or benefit from police presence. The following descriptions
are the request calls included in the original analysis. Only 911 calls
with the given descriptors are considered for this analysis.

```{r, message=F, warning=F, results=F}
# Categories given by PPG Buffalo 

mental_health_desc <- c("LOUD NOISE", "NEIGHBOR TROUBLE","TRESPASSING","CUSTOMER TROUBLE",
"DOMESTIC TROUBLE","CHECK WELFARE","UNWELCOME GUEST","DRUNK","NARCOTICS","MISCELLANEOUS",
"ASSIST CITIZEN","PERSON SOLICITING", "JUVENILE TROUBLE","PROPERTY DISPUTE","MOTORIST STRANDED","LANDLORD TROUBLE","SUICIDE ATTEMPT","PERSON DOWN","PERSON SCREAMING","JUVENILE FOUND","INDECENT EXPOSURE","LABOR DISPUTE","PROSTITUTION","GAMBLING" )

# Create a new dataset with only mental health descriptors
mental_health_data <- calls_sf %>%
  filter( DESCRIPTIO %in% mental_health_desc ) %>%
  rename(
    description = DESCRIPTIO,
    date = DATEREPORT,
    full_address = Buff_remov,
    street_name = STRNAME,
    zipcode = zcta5ce10
  )
```

### 2.3 Analytic Methods

#### 2.3.1 Call Type Analysis

##### Categorize call type, map dominate calls, frequency bar chart

A summary table of each description and the frequency of each
description is created. High call types can indicate areas that may lack
the necessary services to assist in these issues, or limited resources
for the community.

```{r, message=F, warning=F}

# Count each description and order by frequency
summary_table_desc <- mental_health_data %>%
  count(description, sort = FALSE) %>%   # counts and sorts descending by n
  rename(Frequency = n)
```

A frequency bar graph can best visualize what call descriptions occur
frequently. However, this does not show where these calls occur. This
bar graph will be saved in a variable for analysis.

```{r, message=F, warning=F}
# ggplot of the call frequency and description
descr_freq_bar <- ggplot(summary_table_desc, aes(x = factor(description), y = Frequency)) + 
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(
    title = "Frequency of Mental Health Calls in Buffalo, New York",
    caption = "March to April of 2024",
    x = "Call Description",
    y = "Frequency"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

In order to map this data, a new dataset will be created organizing and
summarizing the mental health data and dropping the geometry to perform
some non-spatial operations. The geometry will be added back after to
let us map the top call frequencies by Zipcode.

```{r, message=F, warning=F}
# Top call and description 
top_calls <- mental_health_data %>%
  st_drop_geometry() %>%
  group_by(zipcode, description) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  slice_max(order_by = n, n = 1, with_ties = FALSE)

# Join back for geometry zip data 
top_calls_zip <- zipcode_json %>%
  left_join(top_calls, by = c("zcta5ce10" = "zipcode")) %>%
  select(zcta5ce10, n, description)

```

With the new dataset, a map is created showing the most frequent mental
health 911 calls in each zipcode. It is important to note that this data
is not normalized. The map is then saved to a variable for later
analysis.

```{r, message=F, warning=F}

call_type_map <- ggplot() +
  geom_sf(data = top_calls_zip, aes(fill = description), color = "black") +
  scale_fill_viridis_d(option = "C") +
  labs(
    title = "Most Frequent Mental Health 911 Calls by ZIP Code (Mar–Apr 2024)",
    caption = "City of Buffalo | Data: Erie County 911"
  ) +
  theme_minimal()
```

#### 2.3.2 Call Frequency Analysis

##### Call volume per zip per 1,000 residents, heatmap/choropleth, identify high volume

A summary table of each description and the frequency within each
zipcode is created. High call volume in specific zipcodes can indicate
similar circumstances as having a high call type.

```{r, message=F, warning=F}
# Count each zipcode and order by frequency
summary_table_zip <- mental_health_data %>%
  count(zipcode, sort = TRUE) %>%   # count how many times each ZIP appears
  rename(Frequency = n)
```

A frequency bar graph is created to visualize where these calls occur
frequently. However, this does not show where these calls occur. This
bar graph will be saved in a variable for analysis.

```{r, message=F, warning=F}
# ggplot of the zipcode frequency and description
zipcode_freq_bar <- ggplot(summary_table_zip, aes(x = factor(zipcode), y = Frequency)) + 
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(
    title = "Frequency of Mental Health Calls per Zipcode in Buffalo, New York",
    caption = "March to April of 2024",
    x = "Zipcode",
    y = "Frequency"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

As done previously, the geometry for the summary table will be dropped
in order to perform non-spatial organization. This is done for the
census data as well. Once organized, the geometry is added back while
simultaneously merging the two tables together.

```{r, message=F, warning=F}

# Drop geometry on zip summary data
zip_frequency <- summary_table_zip %>%
  st_drop_geometry() %>%
  group_by(zipcode)

# Drop geometry on census data
census_drop <- census_clean %>%
  st_drop_geometry()

# Merge with census data
zip_data <- zipcode_json %>%
  left_join(zip_frequency, by = c("zcta5ce10" = "zipcode")) %>%
  left_join(census_drop, by = c("geoid10" = "GEOID")) %>%
  mutate(Frequency = replace_na(Frequency, 0))

```

To normalize this data, a new column is created with the frequency /
total population multiplied by 1000. This normalization allows for a
standardized analysis to be done without population bias.

```{r, message=F, warning=F}
# Create normalization data for every 1000 people
zip_data <- zip_data %>%
  mutate(calls_per_1000 = (Frequency / total_pop) * 1000)
```

With this normalized data, a map can be created to show the zip codes in
Buffalo with high call volumes per 1000 people. This map will be saved
to a variable for analysis.

```{r, message=F, warning=F}

# Palette creation for color
pal <- colorNumeric("viridis", domain = zip_data$calls_per_1000)

# Normalize
normalize_map <- leaflet(zip_data) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    fillColor = ~pal(calls_per_1000),
    color = "black",
    weight = 1,
    fillOpacity = 0.8,
    label = ~paste0(
      "<strong>ZIP: </strong>", zcta5ce10, "<br>",
      "<strong>Calls per 1,000: </strong>", round(calls_per_1000, 2), "<br>",
      "<strong>Population: </strong>", total_pop
    ) %>% lapply(htmltools::HTML)
  ) %>%
  addLegend(pal = pal, values = ~calls_per_1000, title = "Calls per 1,000 Residents")
```

#### 2.3.3 Income vs Call Activity Analysis

##### Regression model, map high call/low-income pattern

With the call descriptions and call volume information processed, a
simple regression model can be run to compare how call frequency is
impacted by median household income. To visualize the relationship, this
will be visualized and saved in a variable.

```{r, message=F, warning=F}

# Run simple regression model
model <- lm(Frequency ~ median_income, data = zip_data)
tidy(model)

# Visualize
regression_plot <- ggplot(zip_data, aes(x = median_income, y = Frequency)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Relationship Between 911 Call Volume and Median Household Income",
    x = "Median Household Income ($)",
    y = "Total 911 Calls"
  ) +
  theme_minimal()
```

# 3. Results

\[\~200 words\]

Tables and figures (maps and other graphics) are carefully planned to
convey the results of your analysis. Intense exploration and evidence of
many trials and failures. The author looked at the data in many
different ways before coming to the final presentation of the data.

Show tables, plots, etc. and describe them.

```{r, fig.width=6, fig.height=3, fig.cap="Map of completely random data"}

ggplotly(descr_freq_bar)

call_type_map

ggplotly(zipcode_freq_bar)

normalize_map

regression_plot

```

```{r}

```

### Dygraphs Example

```{r}

```

# 4. Conclusions

\[\~200 words\]

Clear summary adequately describing the results and putting them in
context. Discussion of further questions and ways to continue
investigation.

# 5. References

All sources are cited in a consistent manner
